---
title: 'Statistical Data Mining: Hierarchical Clustering, Silhouette Plots, PCA, K-means Clustering
  with Gap Statistics'
author: "Thankam Abish"
date: "March 18, 2020"
output:
  html_document:
    df_print: paged
---

```{r echo = T, results = 'hide', message=FALSE, warning=FALSE}
library("multtest")
library("fpc")
library("cluster")
library("bootcluster")
```

### US Arrests Data
#### Part (a) Hierarchical Clustering on states with euclidean and complete linkage (3 clusters)
```{r}
data(USArrests)
arrests <- USArrests
set.seed(12353)

distance <- dist(arrests, method = "euclidean")
hc <- hclust(distance, method = "complete")
x11()
plot(hc, hang = -1, labels = row.names(arrests))

rect.hclust(hc, k=3, border="red")
ct <- cutree(hc, h=140)
cluster_1 <- names(ct[ct == 1]) 
cluster_2 <- names(ct[ct == 2]) 
cluster_3 <- names(ct[ct == 3])
```

#### Part (b) Same as (a) but with std dev 1

```{r}
arrests_scaled <- scale(arrests)
distance <- dist(arrests_scaled, method = "euclidean")
hc_scaled <- hclust(distance, method = "complete")
x11()
plot(hc_scaled, hang = -1, labels = row.names(arrests_scaled))
```

### Protein Data
#### Part (a) Plots for different linkages

```{r}
ex2 <- read.csv("Ch10Ex11.csv", header = FALSE)
distance <-  as.dist(1 - cor(ex2))
x11()
plot(hclust(distance, method = "complete"))
x11()
plot(hclust(distance, method = "single"))
x11()
plot(hclust(distance, method = "average"))
```

#### Part (b) Find which genes differ the most by utilizing PCA

```{r}
output <-  prcomp(t(ex2))
load <-  apply(output$rotation, 1, sum)
idx <-  order(abs(load), decreasing = TRUE)
idx[1:10]
```

### Seeds Data: Hierarchical Clustering

```{r}
data <- read.delim("seeds.txt")
data_x <- data[1:7]
distance <- dist(data_x, method = "euclidean")
hc1 <- hclust(distance, method = "complete")
hc2 <- hclust(distance, method = "single")
hc3 <- hclust(distance, method = "average")

store <- c()
k <- c()
for (i in 2:8){
  k <- c(k, i)
  ct <- cutree(hc1, k=i)
  si <- silhouette(ct, dist = distance)
  avg_width <- summary(si)$avg.width
  store <- c(store, avg_width)
}
store

store2 <- c()
for (i in 2:8){
  ct <- cutree(hc2, k=i)
  si <- silhouette(ct, dist = distance)
  avg_width <- summary(si)$avg.width
  store2 <- c(store2, avg_width)
}
store2

store3 <- c()
for (i in 2:8){
  ct <- cutree(hc3, k=i)
  si <- silhouette(ct, dist = distance)
  avg_width <- summary(si)$avg.width
  store3 <- c(store3, avg_width)
}
store3

x11()
plot(k, store, main = 'Average Silhouttes for Different k Values and Linkages',xlab = 'k value', ylab = 'Average Silhouettes',type = 'o', col = 'blue',ylim=c(-0.5,0.7), lty = 1)
points(k, store2, col="purple", pch="x")
lines(k, store2, col="purple", lty = 2)
points(k, store3, col="red", pch="+")
lines(k, store3, col="red", lty = 3)
legend(2, -.15,legend=c("complete","single","average"), col=c("blue","purple","red"),
       pch=c("o","x","+"),lty=c(1,2,3), ncol=1)
```

#### Create 2 clusters

```{r}
## 2 Clusters cut
ct1 <- cutree(hc1, k=2)
si1 <- silhouette(ct1, dist = distance)
ct2 <- cutree(hc2, k=2)
si2 <- silhouette(ct2, dist = distance)
ct3 <- cutree(hc3, k=2)
si3 <- silhouette(ct3, dist = distance)

## Plots with 2 clusters
x11()
plot(si1, col = 2:3)
x11()
plot(si2, col = 2:3)
x11()
plot(si3, col = 2:3)
```

#### Create 3 clusters

```{r}
ct1 <- cutree(hc1, k=3)
si1 <- silhouette(ct1, dist = distance)
ct2 <- cutree(hc2, k=3)
si2 <- silhouette(ct2, dist = distance)
ct3 <- cutree(hc3, k=3)
si3 <- silhouette(ct3, dist = distance)

x11()
plot(si1, col = 2:4)
x11()
plot(si2, col = 2:4)
x11()
plot(si3, col = 2:4)
```

#### Plots of seed data clustering with different linkages

```{r}
idx <- sample(c(1:length(data[,1])), 70)
data <- data[idx, ]
data_x <- data[1:7]
distance <- dist(data_x, method = "euclidean")
hc1 <- hclust(distance, method = "complete")
hc2 <- hclust(distance, method = "single")
hc3 <- hclust(distance, method = "average")

x11()
plot(hc1, hang = -1, labels = data$Seed.Group)
rect.hclust(hc1, k=3, border="red")
x11()
plot(hc2, hang = -1, labels = data$Seed.Group) ## Not great
rect.hclust(hc2, k=2, border="blue")
x11()
plot(hc3, hang = -1, labels = data$Seed.Group)
rect.hclust(hc3, k=3, border="green")

ct1 <- cutree(hc1, k = 3)
ct2 <- cutree(hc2, k = 3)
ct3 <- cutree(hc3, k = 3)

table(ct1, data$Seed.Group)
table(ct2, data$Seed.Group)
table(ct3, data$Seed.Group)
```

### Seed Data: K-means Clustering

```{r}
####  Part (b) Kmeans Clustering

##  gap statistics - kmeans
gap_kmeans <- clusGap(data_x, kmeans, nstart = 20, K.max = 10, B = 100)
x11()
plot(gap_kmeans, main = "Gap Statistic: kmeans")

## kmeans
data_x <- scale(data_x)
km <- kmeans(data_x, centers = 3, nstart = 10)
x11()
plot(data_x[ ,c("Length.Kernel", "Width.Kernel")], col = km$cluster, main = "Example: k-means")
points(km$centers[, c("Length.Kernel", "Width.Kernel")], col = 1:3, pch = 8, cex = 2)

##  how well does the clustering match the labels
table(km$cluster, data$Seed.Group)
length(km$cluster)
length(data$Seed.Group)

```

